Analyzing X-ray images in Python with scikit-image

### 2016-08-21 skimage_tomo.tex
Suggested updates for last section

    \texttt{scikit-image} is an open-source image processing toolkit for
    the Python language, that supports a large variety of file formats
    and is compatible with 2-D and 3-D images. The toolkit exposes a
    simple programming interface, with thematic modules grouping
    functions according to their purpose, such as image restoration,
    segmentation, and measurements. \texttt{scikit-image} users benefit
    from the rich ecosystem of Python for science, with powerful
    visualization libraries or machine learning toolkits.
    \texttt{scikit-image} combines a gentle
    learning curve, versatile image processing capabilities, and the
    scalable performances required for the high-throughput analysis of
    X-ray imaging data.

@@

\texttt{scikit-image} has a simple Application Programming Interface
(API), based almost only on
functions. Most functions take an image (\emph{i.e} a multi-dimensional
array) as input parameter:
\begin{lstlisting}
>>> from skimage import filters
>>> im_gaussian = filters.gaussian(im)
\end{lstlisting}

@@

\begin{figure*}
    \centerline{\includegraphics[width=0.99\textwidth]{tomo_gallery}}
    \caption{\csentence{Typical image processing operations with
	\texttt{scikit-image}.} From left to right and top to bottom.
	Data are synthetic, unless stated otherwise.
	\textbf{Filtering -} Top: non-local means denoising of an image
	with a fine-grained texture, acquired by \emph{in situ}
	synchrotron microtomography during
	glass melting \citep{Gouillart2012}. Bottom: total-variation
	denoising of an image with two phases, corresponding to
	phase-separating silicate melts observed by \emph{in situ
	tomography} \citep{Bouttes2015}.
	\textbf{Feature extraction -} Top: Hubble deep field (NASA,
	public domain), blob detection using the
	Laplacian of Gaussian method. Bottom: ridge detection using the
	leading value of the Hessian matrix. 
	\textbf{Segmentation - } Top: super-pixel segmentation
	of a CT slice of the human head \citep{tomo_wikipedia}, using
	Felzenszwalb's algorithm \citep{Felzenszwalb2004}. Bottom: random
	walker segmentation (right) of noisy image (top-left corner), using
	histogram-determined markers (bottom-left corner).
	\textbf{Measures -} Top: visualization of local diameter
	(color-coded on the skeleton curve) of an
	interconnected phase (represented in violet).  Bottom: particles color-coded according
	to their extent.
\label{fig:tomo_gallery}}
\end{figure*}

@@

Given the large size of tomography datasets, the speed of execution of image
processing operations is a critical issue. \texttt{scikit-image} relies mostly
on calls to \texttt{NumPy} operations, with most of them that are performed in
optimized compiled code (C or Fortran). Performance-critical parts of the code
that cannot call efficient \texttt{NumPy} code are implemented in
\texttt{Cython}. \texttt{Cython} \citep{Behnel2011} is an extension of the
Python language that supports explicit type declarations, and is compiled
directly to C.

However, basic \texttt{scikit-image} code runs on a single core. Computing
workstations and servers used for X-ray imaging typically have several tens of
cores. Parallelization of the computing workflow can be achieved in different
ways. The most trivial parallelization scheme consists in applying the same
workflow to different images, on different cores. However, finer-grained
parallelization is preferable when prototyping the processing workflow.

An easy solution consists in dividing an image into smaller images (with
or without overlap, depending on the operation), and to apply the
same operation on the different sub-images, on different cores. Creating overlapping chunks is easy with the dedicated function \texttt{view\_as\_windows} (or \texttt{view\_as\_blocks} for contiguous non-overlapping chunks):
\begin{lstlisting}[language=Python]
>>> from skimage import util, data
>>> im = data.camera()
>>> im.shape
(512, 512)
>>> # chunks of size 120 with 8-pixel overlap
>>> chunks = util.view_as_windows(im, (128, 128), step=120)
>>> chunks.shape
(4, 4, 128, 128)
\end{lstlisting}
The \texttt{joblib} library enables easy parallel processing. Looping over the different blocks, and dispatching the computation over several cores, is realized with the following syntax:
\begin{lstlisting}
>>> from joblib import Parallel, delayed
>>> filtered_chunks = Parallel(n_jobs=4)(delayed(filters.gaussian)(chunks[i, j]) for i in range(4) for j in range(4))
\end{lstlisting}
\texttt{scikit-image} also offers experimental support for a more integrated parallel processing pipeline, thanks to the \texttt{dask} \citep{dask} module:
\begin{lstlisting}
filtered_im = util.apply_parallel(filters.gaussian, im, depth=8)
\end{lstlisting}
The size of chunks is determined automatically from the number of available cpus, or can be specified by the user.

On a different note, prototyping the image processing workflow requires several trial and error
executions of scripts containing different operations. The \texttt{joblib}
module provides a caching mechanism that avoids the repetition of function
calls, if their arguments have not changed:
\begin{lstlisting}
>>> from skimage import filters, data
>>> from joblib import Memory
>>> mem = Memory(cachedir='/tmp/joblib')
>>> median = mem.cache(filters.median)
>>> im = data.camera()
>>> filtered_im = median(im, np.ones((3, 3)))
__________________________
[Memory] Calling skimage.filters.rank.generic.median...
median(array([[156, ..., 152],
       ..., 
       [121, ..., 111]], dtype=uint8), array([[ 1.,  1.,  1.],
       [ 1.,  1.,  1.],
       [ 1.,  1.,  1.]]))
__________________________median - 0.0s, 0.0min
>>> filtered_again = median(im, np.ones((3, 3)))
>>> # the above call did not trigger an evaluation
\end{lstlisting}

Finally, let us note that the biggest performance improvements often come from
the improvement of algorithms (as opposed to computing architectures only). For
example, non-local means denoising \citep{Buades2005} is a costly operation,
since it requires several nested loops, on all pixels and on neighboring
patches to be compared with the pixel-centered patch. Thanks to the
implementation of a more recent algorithm \citep{Darbon2008} that modifies the
internal organization of loops, it was possible to gain a factor of 5 to 10.
The large size of the \texttt{scikit-image} community ensures that possible
algorithmic improvements are discussed regularly.

@@

The quality of software documentation is an important component to the
success of a package aimed at scientists.
\texttt{scikit-image} users have access to different kinds of
documentation. Functions of the API are all
documented using the same standard \citep{Pawlik2015} as other major
Scientific Python packages. The standard include a description of all
input and output variables and their data type, together with
explanations on what the function does and how to use it. Function
documentation is accessed either online, or within the development
environment (IPython, Spyder, Jupyter Notebook...).

@@

In addition, a popular documentation tool is the graphical gallery of
examples (\nolinkurl{http://scikit-image.org/docs/dev/auto_examples/}), a part of which is displayed in Fig.~\ref{fig:gallery}.
The principle of the gallery is to accelerate the learning curve of image
processing, by showcasing graphical examples of image processing
operations. Such examples are organized as an array of thumbnails with a
short title (see Fig.~\ref{fig:gallery} left). These thumbnails link to
the webpage of the corresponding example, which features a mini-tutorial
on the image processing method, the code needed to run the example and
the figure generated by the example. Since the graphical gallery is an
efficient way to inform users about the features of
\texttt{scikit-image}, every new feature integrated in the package must
include an example for the gallery.
Longer tutorials and a more narrative documentation is available as well
in the online User Guide of \texttt{scikit-image}. The User Guide explains in
particular "big picture" aspects of \texttt{scikit-image}, such as its
use of \texttt{NumPy} arrays as images, or how the package interacts with
other parts of the scientific Python ecosystem.

@@

Estimating the number of active users of an open-source package is a
difficult task. Statistics of downloads, for example, largely
overestimate the number of active users, all the more if the package is
bundled with others in a software distribution, like Anaconda or
Canopy. A view closer to reality can be obtained by analyzing the
statistics of visits of the online help, available on the project
website.  As of the first half of 2016, 20000 unique visitors visit the
website of scikit-image every month \url{http://scikit-image.org/}. These
users come from 80 countries, located almost equally in Europe, America
and Asia (Oceania and Africa representing a much smaller fraction of
visitors).

@@

\texttt{scikit-image} is developed by a diverse team of volunteers, with
more than 170 developers that have contributed to the package. The large
number of developers and users is a key element of the sustainability of
\texttt{scikit-image}. The
development process takes place on the GitHub platform
\url{https://github.com/scikit-image/scikit-image}, where users and
developers propose and discuss new contributions, or report bugs or ideas
for improvements.
A release cycle of one or two releases every year
ensures that new features are proposed to users on a regular basis.

@@

\texttt{scikit-image} offers a wide variety of image processing
algorithms, using a simple interface natively compatible with 2D and 3D
images. It is well integrated into the Scientific Python ecosystem, so
that it interfaces well with visualization libraries or other data
processing packages. \texttt{scikit-image} has seen a tremendous growth
since its creation in 2010, both in terms of users and proposed features.
In addition to the growing number of scientific teams that use
\texttt{scikit-image} for processing X-ray-modalities images,
domain-specific tools are now using \texttt{scikit-image} as a dependency
to build upon. Examples include \texttt{tomopy} \citep{Gursoy2014} for
tomographic reconstruction, or \texttt{DIOPTAS} \citep{Prescher2015} for
the reduction and exploration of X-ray diffraction data. It is likely
that more application-specific softwares will benefit from depending on
\texttt{scikit-image} in the future, since \texttt{scikit-image} strives
to be domain-agnostic and to keep a stable interface. On the end-user
side, future work includes a better integration of parallel processing
capabilities, completion of full 3-D compatibility, an enriched narrative
documentation, and the extension of the number of algorithms.

>>>>

    \texttt{scikit-image} is an open-source image processing toolkit for
    the Python language, that supports a large variety of file formats
    and is compatible with 2-D and 3-D images. The toolkit exposes a
    simple programming interface, corresponding to functions performing
    operations on images, and different thematic modules that group these
    functions according to their purpose, such as image restoration,
    segmentation, measurements, etc. \texttt{scikit-image} users benefit
    from the rich ecosystem of Python for science, with powerful
    visualization libraries or machine learning toolkits.
    \texttt{scikit-image} combines a gentle
    learning curve, versatile image processing capabilities, and the
    scalable performances required for the high-throughput analysis of
    X-ray imaging data.

@@

\texttt{scikit-image} has a very simple Application Programming Interface
(API), based almost only on
functions. Most functions take an image (\emph{i.e} a multi-dimensional
array) as input parameter:
\begin{lstlisting}
>>> from skimage import filters
>>> im_gaussian = filters.gaussian(im)
\end{lstlisting}

@@

\begin{figure*}
    \centerline{\includegraphics[width=0.99\textwidth]{tomo_gallery}}
    \caption{\csentence{Typical image processing operations with
	\texttt{scikit-image}.} From left to right and top to bottom.
	Data are syntheric, unless stated otherwise.
	\textbf{Filtering -} Top: non-local means denoising of an image
	with a fine-grained texture, acquired by \emph{in situ}
	synchrotron microtomography during
	glass melting \citep{Gouillart2012}. Bottom: total-variation
	denoising of an image with two phases, corresponding to
	phase-separating silicate melts observed by \emph{in situ
	tomography} \citep{Bouttes2015}.
	\textbf{Feature extraction -} Top: Hubble deep field (NASA,
	public domain), blob detection using the
	Laplacian of Gaussian method. Bottom: ridge detection using the
	leading value of the Hessian matrix. 
	\textbf{Segmentation - } Top: super-pixel segmentation
	of a CT slice of the human head \citep{tomo_wikipedia}, using
	Felzenszwalb's algorithm \citep{Felzenszwalb2004}. Bottom: random
	walker segmentation (right) of noisy image (top-left corner), using
	histogram-determined markers (bottom-left corner).
	\textbf{Measures -} Top: visualization of local diameter
	(color-coded on the skeleton curve) of an
	interconnected phase (represented in violet).  Bottom: particles color-coded according
	to their extent.
\label{fig:tomo_gallery}}
\end{figure*}

@@

Given the large size of tomography datasets, the execution speed of image
processing operations is of critical concern. \texttt{scikit-image} relies mostly
on calls to \texttt{NumPy} operations, of which most are performed in
optimized compiled code (C or Fortran). Performance-critical parts of \texttt{scikit-image}
that cannot call efficient \texttt{NumPy} code are implemented in
\texttt{Cython}. \texttt{Cython} \citep{Behnel2011} is an extension of the
Python language that supports explicit type declarations, and is compiled
directly to C.

However, basic \texttt{scikit-image} code runs on a single core. Computing
workstations and servers used for X-ray imaging typically have several tens of
cores. Parallelization of the computing workflow can be achieved in multiple
ways. The most trivial parallelization scheme consists of applying the same
workflow to different images, on different cores. However, finer-grained
parallelization is preferable when prototyping the processing workflow.

An easy solution consists in dividing an image into smaller images (with
or without overlap, depending on the operation), and to apply the
same operation on the different sub-images, on different cores. Creating overlapping chunks is easy with the dedicated function \texttt{view\_as\_windows} (or \texttt{view\_as\_blocks} for contiguous non-overlapping chunks):
\begin{lstlisting}[language=Python]
>>> from skimage import util, data
>>> im = data.camera()
>>> im.shape
(512, 512)
>>> # chunks of size 120 with 8-pixel overlap
>>> chunks = util.view_as_windows(im, (128, 128), step=120)
>>> chunks.shape
(4, 4, 128, 128)
\end{lstlisting}
The \texttt{joblib} module enables easy parallel processing. Looping over the different blocks, and dispatching the computation over several cores, is realized with the following syntax:  
\begin{lstlisting}
>>> from joblib import Parallel, delayed
>>> filtered_chunks = Parallel(n_jobs=4)(delayed(filters.gaussian)(chunks[i, j]) for i in range(4) for j in range(4))
\end{lstlisting}
\texttt{scikit-image} also offers experimental support for a more integrated parallel processing pipeline, thanks to the \texttt{dask} \citep{dask} module:
\begin{lstlisting}
filtered_im = util.apply_parallel(filters.gaussian, im, depth=8)
\end{lstlisting}
The size of chunks is determined automatically from the number of available cpus, or can be specified by the user.

Another situation that often arises is that, while prototyping a
workflow, scripts (containing the image processing pipeline) are run
several times to experiment with parameters. The \texttt{joblib}
module provides a caching mechanism that avoids the repetition of function
calls if their arguments have not changed:
\begin{lstlisting}
>>> from skimage import filters, data
>>> from joblib import Memory
>>> mem = Memory(cachedir='/tmp/joblib')
>>> median = mem.cache(filters.median)
>>> im = data.camera()
>>> filtered_im = median(im, np.ones((3, 3)))
__________________________
[Memory] Calling skimage.filters.rank.generic.median...
median(array([[156, ..., 152],
       ..., 
       [121, ..., 111]], dtype=uint8), array([[ 1.,  1.,  1.],
       [ 1.,  1.,  1.],
       [ 1.,  1.,  1.]]))
__________________________median - 0.0s, 0.0min
>>> filtered_again = median(im, np.ones((3, 3)))
>>> # the above call did not trigger another evaluation
\end{lstlisting}

Finally, let us note that the biggest performance improvements often come from
the improvement of algorithms (as opposed to computing architectures only). For
example, non-local means denoising \citep{Buades2005} is a costly operation,
since it requires several nested loops, on all pixels and on neighboring
patches to be compared with the pixel-centered patch. Thanks to the
implementation of a more recent algorithm \citep{Darbon2008} that modifies the
internal organization of loops, it was possible to improve execution
time by a factor of roughly ten times.
The large size of the \texttt{scikit-image} community makes it likely for
algorithmic improvements to be discussed regularly.

@@

The quality of software documentation is (perhaps especially)
important in software aimed at scientists.
\texttt{scikit-image} users have access to several kinds of
documentation. Functions of the API are all
documented using the same standard \citep{Pawlik2015} as other major
Scientific Python packages. The standard include a description of all
input and output variables and their data types, together with
explanations of what each function does and how to use it. Function
documentation is accessed either online, or within the development
environment (IPython, Spyder, Jupyter Notebook...).

@@

In addition, a graphical gallery of examples
(\nolinkurl{http://scikit-image.org/docs/dev/auto_examples/}),
part of which is displayed in Fig.~\ref{fig:gallery},
showcases graphical examples of image processing
operations.  The examples are organized as an array of thumbnails with a
short title (see Fig.~\ref{fig:gallery} left). These thumbnails link to
the webpage of the corresponding example, which features a mini-tutorial
on the image processing method, the code needed to run the example and
the figure generated by the example. Since the graphical gallery is an
efficient way to inform users about the features of
\texttt{scikit-image}, every new feature integrated in the package must
include an example for the gallery.
Longer tutorials and a more narrative documentation is available as well
in the online User Guide of \texttt{scikit-image}. The User Guide explains in
particular "big picture", foundational aspects of \texttt{scikit-image}, such as its
use of \texttt{NumPy} arrays as images, or how the package interacts with
other parts of the scientific Python ecosystem.

@@

Estimating the number of active users of an open-source package is a
difficult task. Download statistics, for example, largely
overestimate the number of active users, all the more if the package is
bundled with others in a software distribution, like Anaconda or
Canopy. A view closer to reality can be obtained by analyzing the
statistics of visits of the online help, available on the project
website.  As of the first half of 2016, 20000 unique visitors visited the
\texttt{scikit-image} website every month at \url{http://scikit-image.org/}. These
users come from 80 countries, located almost equally in Europe, America
and Asia (Oceania and Africa representing a much smaller fraction of
visitors).

@@

\texttt{scikit-image} is developed by a diverse team of volunteers, with
more than 170 developers having contributed to the package. The large
number of developers and users is a key element of the sustainability of
\texttt{scikit-image}. The
development process takes place on the GitHub platform
\url{https://github.com/scikit-image/scikit-image}, where users and
developers propose and discuss new contributions, report bugs or submit ideas
for improvements.
A release cycle of one or two releases every year
ensures that new features are propagated to users on a regular basis.

@@

\texttt{scikit-image} offers a wide variety of image processing
algorithms, using a simple interface natively compatible with 2D and 3D
images. It is well integrated into the Scientific Python ecosystem, so
that it interfaces well with visualization libraries and other data
processing packages. \texttt{scikit-image} has seen tremendous growth
since its creation in 2009, both in terms of users and proposed features.
In addition to the growing number of scientific teams that use
\texttt{scikit-image} for processing images of various X-ray modalities,
domain-specific tools are now using \texttt{scikit-image} as a dependency
to build upon. Examples include \texttt{tomopy} \citep{Gursoy2014} for
tomographic reconstruction, or \texttt{DIOPTAS} \citep{Prescher2015} for
the reduction and exploration of X-ray diffraction data. It is likely
that more application-specific software will benefit by depending on
\texttt{scikit-image} in the future, since \texttt{scikit-image} strives
to be domain-agnostic and to keep the function interface stable. On the end-user
side, future work includes better integration of parallel processing
capabilities, completion of full 3-D compatibility, an enriched narrative
documentation, speed enhancements, and expansion of the set of supported algorithms.
